---
title: "Getting Started with GeoTessera"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with GeoTessera}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The **GeoTessera** package provides an R interface to Tessera geofoundation model
embeddings. Tessera is a foundation model that processes Sentinel-1 and Sentinel-2
satellite imagery to generate 128-channel dense representation maps at 10m resolution.

These embeddings capture rich semantic information about land cover, land use, and
environmental conditions, making them useful for:

- Land cover classification
- Change detection
- Environmental monitoring
- Agricultural applications
- Urban analysis

## Installation

```{r install, eval=FALSE}
# Install from GitHub
remotes::install_github("lassa_sentinel/GeoTessera")

# Or install from local source
install.packages("path/to/GeoTessera", repos = NULL, type = "source")
```

```{r setup}
library(GeoTessera)
```

## Understanding the Data Structure

Tessera data is organized hierarchically:

```
World Coordinates (any lon/lat)
    |
Blocks (5 x 5 degree regions) - for registry organization
    |
Tiles (0.1 x 0.1 degree regions) - individual data files
    |
Pixels (10m resolution) - 1111 x 1111 per tile
    |
Embeddings (128 channels per pixel)
```

Each tile covers approximately 11 km x 11 km and contains about 1.2 million pixels,
each with 128 embedding values.

## Creating a GeoTessera Client

The main entry point is the `geotessera()` function:

```{r client}
# Create a client with default settings
gt <- geotessera()

# Print information about the client
print(gt)
```

You can customize the client with various options:

```{r client-custom}
gt <- geotessera(
  dataset_version = "v1",           # Dataset version
  cache_dir = "~/.cache/geotessera", # Where to cache registry files
  embeddings_dir = "./data",         # Where to store downloaded embeddings
  verify_hashes = TRUE               # Verify SHA256 hashes on downloads
)
```

## Quick Start: Cambridge Region Example

This section uses the Cambridge, UK region as a practical example. This area
contains 4 tiles (approximately 400 MB total) and is a good size for getting started.

### Step 1: Initialize and Query

```{r cambridge-query}
# Create client
gt <- geotessera()

# Cambridge region bounding box (covers 4 tiles)
cambridge_bbox <- c(0.086174, 52.183432, 0.151062, 52.206318)

# Check how many tiles are available
n_tiles <- gt$embeddings_count(
  bbox = list(
    xmin = cambridge_bbox[1],
    ymin = cambridge_bbox[2],
    xmax = cambridge_bbox[3],
    ymax = cambridge_bbox[4]
  ),
  year = 2024
)
print(paste("Tiles available:", n_tiles))
# Expected: 4 tiles
```

### Step 2: Get Tile Metadata

```{r cambridge-tiles}
# Get tile metadata (doesn't download data yet)
tiles <- gt$get_tiles(
  bbox = list(
    xmin = cambridge_bbox[1],
    ymin = cambridge_bbox[2],
    xmax = cambridge_bbox[3],
    ymax = cambridge_bbox[4]
  ),
  year = 2024
)

# View tile information
print(tiles)
# Columns: lon, lat, year, embedding_hash, scales_hash, embedding_size, scales_size

# The 4 tiles should be:
# grid_0.05_52.15, grid_0.05_52.25, grid_0.15_52.15, grid_0.15_52.25
```

### Step 3: Download and Export as GeoTIFFs

```{r cambridge-export}
# Export tiles as GeoTIFFs (downloads data and converts)
output_dir <- "cambridge_embeddings"
paths <- gt$export_embedding_geotiffs(
  tiles = tiles,
  output_dir = output_dir,
  compress = "lzw",
  progress = TRUE
)

# List exported files
print(paths)
```

### Step 4: Load and Inspect a Tile

```{r cambridge-load}
# Download a single tile containing a specific point
tile <- gt$download_tile(
  lon = 0.1,      # A point within Cambridge
  lat = 52.2,
  year = 2024
)

# View tile information
print(tile)
print(tile$get_grid_name())   # "grid_0.05_52.15"
print(tile$get_bounds())      # xmin=0.0, ymin=52.1, xmax=0.1, ymax=52.2

# Load the embedding array
embedding <- tile$load_embedding()
print(dim(embedding))
# [1] 1111 1111  128 (height, width, channels)

# Access a specific pixel's embedding (center of tile)
pixel_embedding <- embedding[556, 556, ]
print(length(pixel_embedding))  # 128 values
```

### Step 5: Sample at Specific Points

```{r cambridge-sample}
# Define sample points within Cambridge
sample_points <- data.frame(
  lon = c(0.10, 0.12, 0.14),
  lat = c(52.19, 52.20, 52.21),
  site_name = c("West Cambridge", "Central", "East Cambridge")
)

# Sample embeddings at these points
samples <- gt$sample_embeddings_at_points(
  points = sample_points,
  year = 2024,
  download = TRUE,
  progress = TRUE
)

# View results
print(dim(samples))
# [1]   3 130 (3 points x (lon + lat + 128 embeddings))

print(names(samples)[1:10])
# "lon" "lat" "embedding_1" "embedding_2" ...
```

## Working with Different Regions

### UK Region (16 tiles)

A larger example covering London:

```{r uk-region}
# UK region around London (16 tiles)
uk_bbox <- list(xmin = -0.1, ymin = 51.3, xmax = 0.1, ymax = 51.5)

# Check tile count
n_tiles <- gt$embeddings_count(bbox = uk_bbox, year = 2024)
print(paste("UK region tiles:", n_tiles))
# Expected: 16 tiles

# Get tile metadata (dry run - no download)
uk_tiles <- gt$get_tiles(bbox = uk_bbox, year = 2024)
print(nrow(uk_tiles))  # 16
```

### Single Tile from a Point

```{r single-point}
# Any point resolves to a specific tile
# Point (0.17, 52.23) -> tile grid_0.15_52.25
tile_coords <- tile_from_world(0.17, 52.23)
print(paste("Tile:", tile_coords$tile_lon, tile_coords$tile_lat))
# "Tile: 0.15 52.25"

# Download just that tile
single_tile <- gt$download_tile(lon = 0.17, lat = 52.23, year = 2024)
print(single_tile$get_grid_name())  # "grid_0.15_52.25"
```

## Exploring Available Data

```{r explore}
# Check available years
years <- gt$registry$get_available_years()
print(years)
# [1] 2017 2018 2019 2020 2021 2022 2023 2024 2025

# Get tile counts by year
counts <- gt$registry$get_tile_counts_by_year()
print(counts)
```

## Different Bounding Box Formats

```{r bbox-formats}
# As a named vector
bbox <- c(xmin = 0.086174, ymin = 52.183432, xmax = 0.151062, ymax = 52.206318)

# As a list
bbox <- list(xmin = 0.086174, ymin = 52.183432, xmax = 0.151062, ymax = 52.206318)

# From an sf object
library(sf)
cambridge <- st_point(c(0.12, 52.2)) |>
  st_sfc(crs = 4326) |>
  st_buffer(0.05)
bbox <- st_bbox(cambridge)

# All formats work with get_tiles()
tiles <- gt$get_tiles(bbox = bbox, year = 2024)
```

## Exporting Specific Bands

```{r export-bands}
# Export only first 3 bands (useful for RGB visualization)
paths <- gt$export_embedding_geotiffs(
  tiles = tiles,
  output_dir = "cambridge_3bands",
  bands = 1:3,
  compress = "lzw"
)
```

## Working with terra

The exported GeoTIFFs work seamlessly with the `terra` package:

```{r terra-integration}
library(terra)

# Load an exported tile
r <- rast("cambridge_embeddings/grid_0.05_52.15.tif")

# Basic information
print(r)
print(nlyr(r))      # 128 layers
print(res(r))       # Resolution (in UTM meters)
print(ext(r))       # Extent (in UTM coordinates)
print(crs(r))       # CRS (UTM zone 31N for Cambridge)

# Extract specific bands
bands_subset <- r[[1:3]]

# Plot a single band
plot(r[[1]], main = "Band 1")

# Create RGB plot from bands 30, 60, 90
plotRGB(r[[c(30, 60, 90)]], stretch = "lin")

# Calculate statistics
global(r[[1]], fun = "mean", na.rm = TRUE)
```

## Applying PCA for Visualization

Reduce 128 bands to 3 principal components for visualization:

```{r pca}
# Load embedding from a tile
tile <- gt$download_tile(lon = 0.1, lat = 52.2, year = 2024)
embedding <- tile$load_embedding()

# Apply PCA (128 -> 3 components)
pca_result <- gt$apply_pca_to_embeddings(
  embeddings = embedding,
  n_components = 3
)

print(dim(pca_result))
# [1] 1111 1111    3

# Export tiles with PCA applied
pca_paths <- gt$export_pca_geotiffs(
  tiles = tiles,
  output_dir = "cambridge_pca",
  n_components = 3
)
```

## Merging Tiles into Mosaics

Combine multiple tiles into a single raster:

```{r merge-mosaic}
# Merge all GeoTIFFs in a directory
gt$merge_geotiffs_to_mosaic(
  input_dir = "cambridge_embeddings",
  output_path = "cambridge_mosaic.tif"
)

# Or use terra directly for more control
library(terra)
tiffs <- list.files("cambridge_embeddings", pattern = "\\.tif$", full.names = TRUE)
rasters <- lapply(tiffs, rast)
mosaic <- do.call(merge, rasters)
writeRaster(mosaic, "cambridge_mosaic_terra.tif")
```

## Discovering Local Tiles

If you have previously downloaded tiles, you can discover them:

```{r discover}
# Auto-detect tile format and find all tiles
tiles <- discover_tiles("cambridge_embeddings")
print(length(tiles))

# Get tiles by format
formats <- discover_formats("./data")
print(names(formats))  # "npy", "geotiff", "zarr"
print(length(formats$geotiff))
```

## Configuration Options

### Cache Directory

```{r cache-config}
# Check default cache location
print(get_cache_dir())

# Set custom cache via environment variable
Sys.setenv(GEOTESSERA_CACHE_DIR = "/path/to/cache")

# Or pass directly to client
gt <- geotessera(cache_dir = "/custom/cache")
```

### Hash Verification

```{r hash-config}
# Disable hash verification for faster downloads (not recommended for production)
gt <- geotessera(verify_hashes = FALSE)
```

## Best Practices

1. **Start with Cambridge**: Use the Cambridge region (4 tiles, ~400 MB) for testing.

2. **Use compression**: Always use `compress = "lzw"` for GeoTIFF exports to
   save disk space (typically 50-70% reduction).

3. **Work regionally**: Process data in chunks rather than loading entire
   countries at once.

4. **Keep hash verification**: Leave `verify_hashes = TRUE` to ensure data
   integrity, especially for archival purposes.

5. **Use point sampling**: For machine learning tasks, sample at points rather
   than downloading entire tiles when possible.

6. **Monitor disk space**: Each tile is approximately 100 MB for NPY format.
   The Cambridge region is ~400 MB; the UK London region (16 tiles) is ~1.6 GB.

## Troubleshooting

### Connection Issues

```{r troubleshoot-connection, eval=FALSE}
# Check if the registry is accessible
tryCatch({
  gt <- geotessera()
  years <- gt$registry$get_available_years()
  print(paste("Connection successful. Years:", paste(years, collapse = ", ")))
}, error = function(e) {
  print(paste("Connection failed:", e$message))
})
```

### Missing Tiles

```{r troubleshoot-tiles, eval=FALSE}
# Some regions may not have coverage
bbox <- list(xmin = -180, ymin = -90, xmax = -170, ymax = -80)  # Antarctica
tiles <- gt$get_tiles(bbox = bbox, year = 2024)
if (nrow(tiles) == 0) {
  message("No tiles available for this region")
}
```

## Next Steps

- See `vignette("machine-learning", package = "GeoTessera")` for ML workflows
- Explore the [Tessera documentation](https://geotessera.org) for more details
- Check the package reference: `help(package = "GeoTessera")`
