---
title: "Summarizing Embeddings for Administrative Regions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summarizing Embeddings for Administrative Regions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
# Most chunks require network access and are set eval=FALSE individually
```

## Introduction

When working with large geographic regions, you often need summary statistics rather
than full pixel-level embeddings. The `summarize_region()` method provides a pipeline
that:

1. Downloads embeddings to a temporary location
2. Optionally masks pixels to match irregular region boundaries
3. Computes user-specified summary statistics
4. Automatically cleans up downloaded data

This vignette demonstrates how to use this functionality with administrative
boundaries from a shapefile.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(GeoTessera)
library(sf)
library(ggplot2)
```

## Demonstration with Simulated Data

Before working with real data, let's demonstrate the summarization workflow
with simulated embeddings and regions.

### Creating Example Regions

```{r demo-regions, fig.width=6, fig.height=5}
# Create 4 example polygon regions (simulating LGAs)
region1 <- st_polygon(list(matrix(c(
  0, 0, 1, 1, 0,
  0, 1, 1, 0, 0
), ncol = 2))) |> st_sfc(crs = 4326)

region2 <- st_polygon(list(matrix(c(
  1, 1, 2, 2, 1,
  0, 1, 1, 0, 0
), ncol = 2))) |> st_sfc(crs = 4326)

region3 <- st_polygon(list(matrix(c(
  0, 0, 1, 1, 0,
  1, 2, 2, 1, 1
), ncol = 2))) |> st_sfc(crs = 4326)

region4 <- st_polygon(list(matrix(c(
  1, 1, 2, 2, 1,
  1, 2, 2, 1, 1
), ncol = 2))) |> st_sfc(crs = 4326)

# Combine into sf object
regions <- st_sf(
  name = c("Region A", "Region B", "Region C", "Region D"),
  land_type = c("urban", "agricultural", "forest", "mixed"),
  geometry = c(region1, region2, region3, region4)
)

# Plot the regions
ggplot(regions) +
  geom_sf(aes(fill = land_type), alpha = 0.7) +
  geom_sf_text(aes(label = name), size = 4) +
  scale_fill_manual(values = c("urban" = "#E41A1C",
                                "agricultural" = "#FFD92F",
                                "forest" = "#4DAF4A",
                                "mixed" = "#984EA3")) +
  labs(title = "Example Administrative Regions",
       subtitle = "Simulating 4 LGAs with different land cover types") +
  theme_minimal()
```

### Simulating Region Embeddings

```{r simulate-region-embeddings}
set.seed(42)

# Simulate mean embeddings for each region type
# Different land types have distinct embedding signatures
simulate_embedding <- function(land_type) {
  base <- switch(land_type,
    "urban" = rnorm(128, mean = 0.5, sd = 0.1),
    "agricultural" = rnorm(128, mean = 0.0, sd = 0.15),
    "forest" = rnorm(128, mean = -0.4, sd = 0.12),
    "mixed" = rnorm(128, mean = 0.1, sd = 0.2)
  )
  base
}

# Generate embeddings for each region
region_embeddings <- lapply(regions$land_type, simulate_embedding)
names(region_embeddings) <- regions$name

# Create embedding matrix
emb_matrix <- do.call(rbind, region_embeddings)
rownames(emb_matrix) <- regions$name

cat("Embedding matrix dimensions:", dim(emb_matrix), "\n")
cat("Rows:", rownames(emb_matrix), "\n")
```

### Region Similarity Heatmap

```{r similarity-heatmap, fig.width=6, fig.height=5}
# Calculate cosine similarity between regions
cosine_sim <- function(a, b) {
  sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2)))
}

n_regions <- nrow(emb_matrix)
sim_matrix <- matrix(0, n_regions, n_regions)
for (i in 1:n_regions) {
  for (j in 1:n_regions) {
    sim_matrix[i, j] <- cosine_sim(emb_matrix[i, ], emb_matrix[j, ])
  }
}
rownames(sim_matrix) <- colnames(sim_matrix) <- regions$name

# Convert to data frame for plotting
sim_df <- as.data.frame(as.table(sim_matrix))
names(sim_df) <- c("Region1", "Region2", "Similarity")

ggplot(sim_df, aes(x = Region1, y = Region2, fill = Similarity)) +
  geom_tile() +
  geom_text(aes(label = round(Similarity, 2)), color = "white", size = 4) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0.5) +
  labs(title = "Region Similarity Matrix",
       subtitle = "Cosine similarity based on mean embeddings") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Embedding Profile Comparison

```{r embedding-profiles, fig.width=7, fig.height=4}
# Compare first 20 dimensions across regions
profile_df <- data.frame(
  Dimension = rep(1:20, 4),
  Value = c(emb_matrix[1, 1:20], emb_matrix[2, 1:20],
            emb_matrix[3, 1:20], emb_matrix[4, 1:20]),
  Region = rep(regions$name, each = 20)
)

ggplot(profile_df, aes(x = Dimension, y = Value, color = Region)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Embedding Profiles by Region",
       subtitle = "First 20 of 128 dimensions shown",
       x = "Embedding Dimension", y = "Value") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Hierarchical Clustering of Regions

```{r region-clustering, fig.width=6, fig.height=4}
# Cluster regions by embedding similarity
dist_matrix <- dist(emb_matrix)
hc <- hclust(dist_matrix, method = "ward.D2")

# Simple dendrogram plot
plot(hc, main = "Hierarchical Clustering of Regions",
     sub = "Based on 128-dimensional embeddings",
     xlab = "", ylab = "Distance")
```

### Map with Cluster Assignment

```{r cluster-map, fig.width=6, fig.height=5}
# Cut tree into 2 clusters
regions$cluster <- factor(cutree(hc, k = 2))

ggplot(regions) +
  geom_sf(aes(fill = cluster), alpha = 0.7) +
  geom_sf_text(aes(label = paste0(name, "\n(", land_type, ")")), size = 3) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Regions Clustered by Embedding Similarity",
       subtitle = "Regions with similar land cover cluster together") +
  theme_minimal()
```

## Loading Administrative Boundaries

We'll use Nigeria's Local Government Areas (LGAs) as an example. This shapefile
contains 774 administrative units at the admin2 level.

```{r load-shapefile, eval=FALSE}
# Load the shapefile
nigeria_lgas <- st_read("shapefiles/nigeria_admin2_shapefile.shp", quiet = TRUE)

# Examine the structure
print(nigeria_lgas[1:5, c("adminCode", "adminName")])
```

```
Simple feature collection with 5 features and 2 fields
  adminCode      adminName                       geometry
1  NG001001      Aba North MULTIPOLYGON (((7.401109 5....
2  NG001002      Aba South MULTIPOLYGON (((7.387495 5....
3  NG015001          Abaji MULTIPOLYGON (((7.045872 9....
4  NG003001           Abak MULTIPOLYGON (((7.811244 5....
5  NG011001      Abakaliki MULTIPOLYGON (((8.4109 6.28...
```

## Summarizing a Single LGA

Let's compute embedding summaries for Aba North, a small LGA in southeastern Nigeria.

```{r select-lga, eval=FALSE}
# Select Aba North LGA
aba_north <- nigeria_lgas[nigeria_lgas$adminName == "Aba North", ]

# View the region
plot(st_geometry(aba_north), main = "Aba North LGA", col = "lightblue")
```

### Basic Summary

The `summarize_region()` method accepts an sf object directly. It will:

- Use the bounding box to determine which tiles to download
- Mask pixels outside the polygon boundary
- Compute summary statistics

```{r basic-summary, eval=FALSE}
# Create GeoTessera client
gt <- geotessera()

# Compute summaries for Aba North
result <- gt$summarize_region(
  region = aba_north,
  year = 2024
)

# View the structure
names(result)
# [1] "summaries" "metadata"

# View metadata
result$metadata
# $region_bbox
# [1]  7.331234  5.081948  7.402708  5.134542
#
# $year
# [1] 2024
#
# $n_tiles
# [1] 4
#
# $is_masked
# [1] TRUE
```

### Default Summary Statistics

By default, `summarize_region()` computes three summaries:

```{r default-summaries, eval=FALSE}
# Mean embedding (128-dimensional vector)
mean_embedding <- result$summaries$mean
length(mean_embedding)
# [1] 128

# First 10 dimensions
head(mean_embedding, 10)

# Centroid embedding (sampled at geographic center)
centroid_embedding <- result$summaries$centroid
length(centroid_embedding)
# [1] 128

# Coverage statistics
result$summaries$coverage
# n_tiles    n_valid_pixels n_total_pixels coverage_percent
#       4           xxxxx          xxxxx            xx.xx
```

## Custom Summary Functions

You can specify which summary functions to use, including custom ones.

### Built-in Summary Functions

GeoTessera provides several built-in summary functions:

| Function | Description |
|----------|-------------|
| `summary_mean` | Mean across all pixels |
| `summary_median` | Median across all pixels |
| `summary_sd` | Standard deviation per dimension |
| `summary_centroid` | Sample at geographic centroid |
| `summary_pixel_count` | Count of valid/total pixels |
| `summary_coverage` | Coverage statistics |
| `summary_quantile(probs)` | Factory for quantile summaries |
| `summary_random_sample(n, seed)` | Factory for random sampling |

```{r builtin-summaries, eval=FALSE}
# Use specific summary functions
result <- gt$summarize_region(
  region = aba_north,
  year = 2024,
  summary_fns = list(
    mean = summary_mean,
    median = summary_median,
    sd = summary_sd,
    coverage = summary_coverage
  )
)

# Standard deviation tells us about embedding variability
sd_embedding <- result$summaries$sd
summary(sd_embedding)
```

### Quantile Summaries

Use `summary_quantile()` to create a function that computes specific quantiles:

```{r quantile-summary, eval=FALSE}
# Create a function for 10th and 90th percentiles
result <- gt$summarize_region(
  region = aba_north,
  year = 2024,
  summary_fns = list(
    q10_90 = summary_quantile(c(0.10, 0.90))
  )
)

# Returns 256 values (128 dimensions × 2 quantiles)
length(result$summaries$q10_90)
```

### Random Sampling

Use `summary_random_sample()` to extract random pixel embeddings:

```{r random-sample, eval=FALSE}
# Sample 50 random pixels (reproducible with seed)
result <- gt$summarize_region(
  region = aba_north,
  year = 2024,
  summary_fns = list(
    samples = summary_random_sample(n = 50, seed = 42)
  )
)

# Returns a matrix: 50 samples × 128 dimensions
dim(result$summaries$samples)
# [1]  50 128
```

### Custom Summary Functions

Create your own summary function. It should accept:

- `embeddings`: List of 3D arrays (height × width × 128)
- `region`: The input region (sf object or bbox)
- `tiles_df`: Tile metadata data frame

```{r custom-summary, eval=FALSE}
# Custom function: compute the embedding at the pixel with maximum brightness
# (using sum across channels as a proxy for "brightness")
summary_max_brightness <- function(embeddings, region = NULL, tiles_df = NULL) {
  max_sum <- -Inf
  max_embedding <- NULL

  for (emb in embeddings) {
    # Sum across channels for each pixel
    pixel_sums <- apply(emb, c(1, 2), sum, na.rm = TRUE)

    # Find max
    max_idx <- which.max(pixel_sums)
    if (pixel_sums[max_idx] > max_sum) {
      max_sum <- pixel_sums[max_idx]
      # Convert linear index to array indices
      idx <- arrayInd(max_idx, dim(pixel_sums))
      max_embedding <- emb[idx[1], idx[2], ]
    }
  }

  names(max_embedding) <- paste0("dim_", seq_along(max_embedding))
  max_embedding
}

# Use the custom function
result <- gt$summarize_region(
  region = aba_north,
  year = 2024,
  summary_fns = list(
    mean = summary_mean,
    max_brightness = summary_max_brightness
  )
)
```

## Processing Multiple Regions

When processing multiple regions, adjacent areas often share tiles. GeoTessera
provides an optimized `summarize_regions_streaming()` function that:

1. **Builds a region-to-tile mapping** - identifies which tiles each region needs
2. **Detects shared tiles** - finds tiles used by multiple regions
3. **Optimizes download order** - processes most-shared tiles first
4. **Downloads each tile only once** - even if multiple regions need it
5. **Uses streaming statistics** - constant memory regardless of region size

### Efficient Multi-Region Processing

```{r multi-region-streaming, eval=FALSE}
# Select adjacent LGAs that share tiles
selected_lgas <- c("Aba North", "Aba South", "Osisioma Ngwa", "Ugwunagbo")
lgas_subset <- nigeria_lgas[nigeria_lgas$adminName %in% selected_lgas, ]

# Process all regions efficiently
result <- summarize_regions_streaming(

  gt = gt,
  regions = lgas_subset,
  year = 2024,
  region_ids = lgas_subset$adminName,
  sample_rate = 0.1,      # Sample 10% of pixels for speed
  mask_to_region = TRUE   # Only include pixels inside each polygon
)

# View optimization statistics
result$metadata
# $n_regions
# [1] 4
#
# $n_unique_tiles
# [1] 11
#
# $tiles_downloaded
# [1] 11
#
# (4 tiles were shared between regions - downloaded only once!)
```

The function returns:

- `summaries`: Named list of mean embeddings (128-dim vector per region)
- `pixel_counts`: Number of pixels processed per region
- `metadata`: Processing statistics

```{r multi-region-results, eval=FALSE}
# Access results for a specific region
aba_north_embedding <- result$summaries[["Aba North"]]
length(aba_north_embedding)
# [1] 128

# View pixel counts
result$pixel_counts
#     Aba North     Aba South Osisioma Ngwa     Ugwunagbo
#         18800         43538        175735        126769

# Combine into a matrix for analysis
embedding_matrix <- do.call(rbind, result$summaries)
dim(embedding_matrix)
# [1]   4 128
```

### Comparison: Sequential vs Optimized

For adjacent regions, the optimized approach is significantly faster:
```
Sequential processing (4 LGAs separately):
  - Each LGA downloads its own tiles
  - Shared tiles downloaded multiple times
  - Total: ~20+ tile downloads

Optimized streaming (summarize_regions_streaming):
  - Builds tile sharing graph
  - Downloads each unique tile once
  - Total: 11 tile downloads (4 shared)
```

### Large-Scale Processing

For processing many regions (e.g., all LGAs in a country), use sampling
to reduce processing time:

```{r large-scale, eval=FALSE}
# Process all LGAs in a state
abia_lgas <- nigeria_lgas[grepl("^NG001", nigeria_lgas$adminCode), ]

result <- summarize_regions_streaming(
  gt = gt,
  regions = abia_lgas,
  year = 2024,
  region_ids = abia_lgas$adminName,
  sample_rate = 0.05,     # 5% sampling for very large jobs
  mask_to_region = TRUE
)
```

### Creating a Summary Data Frame

Combine embeddings with administrative metadata:

```{r summary-dataframe, eval=FALSE}
# Create data frame with LGA info and embeddings
embedding_matrix <- do.call(rbind, result$summaries)
lga_embeddings <- data.frame(
  adminName = names(result$summaries),
  embedding_matrix
)

# Add area information
lga_embeddings$area_km2 <- as.numeric(st_area(lgas_subset)) / 1e6

# Now you can use this for clustering, classification, etc.
head(lga_embeddings[, 1:5])
```

## Controlling Masking Behavior

By default, pixels outside the region boundary are masked (set to NA). You can
disable this to include all pixels within the bounding box:

```{r no-masking, eval=FALSE}
# Without masking - uses rectangular bounding box
result_unmasked <- gt$summarize_region(
  region = aba_north,
  year = 2024,
  mask_to_region = FALSE
)

# With masking (default) - only pixels inside the polygon
result_masked <- gt$summarize_region(
  region = aba_north,
  year = 2024,
  mask_to_region = TRUE
)

# Compare coverage
result_unmasked$summaries$coverage["n_valid_pixels"]
result_masked$summaries$coverage["n_valid_pixels"]
# Masked version will have fewer valid pixels
```

## Using Bounding Boxes

You can also use simple bounding boxes instead of sf objects:

```{r bbox-summary, eval=FALSE}
# Define a bounding box (xmin, ymin, xmax, ymax)
bbox <- c(7.33, 5.08, 7.40, 5.14)

result <- gt$summarize_region(
  region = bbox,
  year = 2024
)

# Note: masking is not applied for bounding boxes
result$metadata$is_masked
# [1] FALSE
```

## Performance Considerations

### Temporary File Management

Downloaded tiles are stored in a temporary directory and automatically cleaned
up when `summarize_region()` completes. If you need to keep the tiles:

```{r keep-tiles, eval=FALSE}
result <- gt$summarize_region(
  region = aba_north,
  year = 2024,
  keep_tiles = TRUE
)

# Access the downloaded Tile objects
tiles <- result$tiles
length(tiles)

# Load embedding from first tile
emb <- tiles[[1]]$load_embedding()
dim(emb)
```

### Large Regions (Memory-Efficient Streaming)

For very large regions (many tiles), the standard `summarize_region()` may run
out of memory because it loads all embeddings before computing statistics.

Use `summarize_region_streaming()` for memory-efficient processing of large
single regions:

```{r large-region-streaming, eval=FALSE}
# Abaji LGA requires 36 tiles - too large for standard approach
abaji <- nigeria_lgas[nigeria_lgas$adminName == "Abaji", ]

# Use streaming approach - processes one tile at a time
result <- gt$summarize_region_streaming(
 region = abaji,
  year = 2024,
  sample_rate = 0.1,       # Sample 10% of pixels
  mask_to_region = FALSE   # Skip masking for speed
)

# Results
result$mean          # 128-dimensional mean embedding
result$n_pixels      # Number of pixels processed
result$metadata$processing_time_secs
```

The streaming approach:

- **Constant memory**: Processes one tile at a time using Welford's online algorithm
- **Sampling support**: Reduce processing time with `sample_rate < 1.0`
- **No memory issues**: Successfully handles regions with 30+ tiles

### Alternative: Point Sampling Within Polygon

Another memory-efficient approach is to sample random points within the
polygon and fetch their embeddings:

```{r point-sampling, eval=FALSE}
# Sample 1000 random points inside the polygon
set.seed(42)
sample_points <- st_sample(abaji, size = 1000)
coords <- st_coordinates(sample_points)
points_df <- data.frame(lon = coords[, 1], lat = coords[, 2])

# Fetch embeddings at those points
embeddings_df <- gt$sample_embeddings_at_points(points_df, year = 2024)

# Compute mean (excluding lon/lat columns)
mean_embedding <- colMeans(embeddings_df[, -(1:2)], na.rm = TRUE)
```

## Example: Clustering LGAs by Land Cover Similarity

Here's a complete example that clusters Nigerian LGAs based on embedding similarity,
using the optimized multi-region streaming approach:

```{r clustering-example, eval=FALSE}
library(GeoTessera)
library(sf)

# Load shapefile
nigeria_lgas <- st_read("shapefiles/nigeria_admin2_shapefile.shp", quiet = TRUE)

# Calculate areas and select small LGAs for demo
nigeria_lgas$area_km2 <- as.numeric(st_area(nigeria_lgas)) / 1e6
small_lgas <- nigeria_lgas[nigeria_lgas$area_km2 < 100, ]
small_lgas <- small_lgas[1:20, ]  # Take first 20 for demo

# Initialize client
gt <- geotessera()

# Compute mean embeddings for all LGAs efficiently
# This optimizes tile downloads when regions share tiles
result <- summarize_regions_streaming(
  gt = gt,
  regions = small_lgas,
  year = 2024,
  region_ids = small_lgas$adminName,
  sample_rate = 0.1,
  mask_to_region = TRUE
)

# Create embedding matrix from results
emb_matrix <- do.call(rbind, result$summaries)

# Hierarchical clustering
dist_matrix <- dist(emb_matrix)
hc <- hclust(dist_matrix, method = "ward.D2")

# Plot dendrogram
plot(hc, main = "LGA Clustering by Embedding Similarity",
     xlab = "", sub = "")

# Cut into clusters
clusters <- cutree(hc, k = 4)

# Add cluster assignments back to spatial data
small_lgas$cluster <- clusters[small_lgas$adminName]

# Plot clusters on map
plot(small_lgas["cluster"], main = "LGA Clusters")
```

## Summary

The `summarize_region()` method provides a convenient pipeline for:

- **Downloading** embeddings for arbitrary regions
- **Masking** to irregular boundaries from shapefiles
- **Computing** flexible summary statistics
- **Cleaning up** temporary files automatically

This enables analysis workflows where you need region-level embeddings rather than
full pixel-level data, such as:

- Administrative unit characterization
- Regional clustering and classification
- Time series analysis at the regional level
- Feature extraction for machine learning models
